---
title: "Sapflow_synoptic_creating_DF"
author: "Roberta Peixoto"
format: html
editor: visual
---

#Getting L2 data - sap + climavue for  vpd

```{r}
setwd("D:/COMPASS_L2-v2-1/v2-1")

SITE <- c("OWC","PTR","CRC",
          "GCW", "GWI", "MSM", "SWH")

VARIABLE <- c("wx-tempavg15",
              "wx-par-den15", #µmol/m2/s Average photosynthetically active radiation (PAR) flux density ; site specific correction factor applied
              "wx-rh15",        #degC     Average temperature of humidity sensor 
              "wx-vpd15",       #kPa      Vapor pressure deficit
              "sapflow-3.5cm",  #mV       Raw sap flow at 3.5 cm depth
              "sapflow-5cm",    #mV       Raw sap flow at 5 cm depth
              "Value_MAC")

all_files <- list.files(
  "./",
  recursive = TRUE,
  full.names = TRUE
)

files <- all_files[
  grepl(paste(SITE, collapse = "|"), basename(all_files)) &
    grepl(paste(VARIABLE, collapse = "|"), basename(all_files)) &
    grepl("_L2_v2-", basename(all_files)) &
    grepl("\\.parquet$", basename(all_files))
]

dat <- bind_rows(
  lapply(files, function(f) {
    message("Reading ", basename(f))
    read_parquet(f)
  })
)


# (1) filtering data for the time period I want to work if nedeed
#dat %>%
#  filter(TIMESTAMP > "2022-07-01 00:00:00",
#    TIMESTAMP < "2023-10-21 00:00:00") ->dat_csf

#(2) averaging per sensor replicates per TIMESTAMP

df_agg <- dat %>%
  group_by(Site, Plot, research_name, Sensor_ID, TIMESTAMP) %>%
  summarise(
    Value_MAC = mean(Value_MAC, na.rm = TRUE),
    .groups = "drop"
  )

#(3) creating our column style df
df_pivoted <- df_agg %>%
  pivot_wider(
    names_from = research_name,
    values_from = Value_MAC
  )


```

#gaps diagnosis

```{r}

library(dplyr)
library(tidyr)
library(lubridate)

# renaming columns
diag_gaps <- df_pivoted %>%
    mutate(year_month = floor_date(TIMESTAMP, "month")) %>%
    group_by(Site, Plot, year_month) %>%
    summarise(
        n_total       = n(),
        n_sapflow     = sum(!is.na(`sapflow-3.5cm`)),
        n_temp        = sum(!is.na(`wx-tempavg15`)),
        n_rh          = sum(!is.na(`wx-rh15`)),
        pct_sapflow   = round(100 * n_sapflow / n_total, 1),
        pct_temp      = round(100 * n_temp    / n_total, 1),
        pct_rh        = round(100 * n_rh      / n_total, 1),
        .groups = "drop"
    ) %>%
    arrange(Site, Plot, year_month)

# Where is therre sapflow but no wx?
diag_gaps <- diag_gaps %>%
    mutate(
        gap_type = case_when(
            pct_sapflow > 0 & pct_temp == 0          ~ "sapflow_only",
            pct_sapflow == 0 & pct_temp > 0          ~ "wx_only",
            pct_sapflow == 0 & pct_temp == 0         ~ "both_missing",
            TRUE                                     ~ "both_present"
        )
    )

# months with problems
diag_gaps %>%
    filter(gap_type != "both_present") %>%
    print(n = Inf)

library(ggplot2)

diag_gaps %>%
  pivot_longer(cols = starts_with("pct_"), names_to = "variable", values_to = "pct_coverage") %>%
  ggplot(aes(x = year_month, y = interaction(Site, Plot), fill = pct_coverage)) +
  geom_tile() +
  facet_wrap(~variable) +
  scale_fill_gradient(low = "white", high = "darkred", limits = c(0,100)) +
  theme_minimal() +
  labs(title = "Data covered (%) per Site/Plot/Month", y = NULL, x = NULL)
```

#df_setup - merge sap and climavue+gw+teros

```{r}

library(dplyr)

# splitting sapflow and climavue
df_sap <- df_pivoted %>%
  select(Site, Plot, Sensor_ID, TIMESTAMP, `sapflow-3.5cm`, `sapflow-5cm`)

df_wx <- df_pivoted %>%
  filter(Plot == "W") %>%
  select(Site, TIMESTAMP, `wx-par-den15`, `wx-rh15`, `wx-tempavg15`, `wx-vpd15`) %>%
  distinct(Site, TIMESTAMP, .keep_all = TRUE)  

# check for timestamp duplication
df_wx %>%
  count(Site, TIMESTAMP) %>%
  filter(n > 1)


# merge climavue and sap per Site+Timestamp
df_merged <- df_sap %>%
  left_join(df_wx, by = c("Site", "TIMESTAMP"))

# plot checks
str(df_merged)
df_merged %>%
  group_by(Plot) %>%
  summarise(
    n = n(),
    pct_temp = round(100 * sum(!is.na(`wx-tempavg15`)) / n, 1),
    pct_rh   = round(100 * sum(!is.na(`wx-rh15`)) / n, 1)
  )

write.csv(df_merged, "sap_temp_rh.csv", row.names = FALSE)
```

#MERGE L2 gw teros wx sensor dataset for stats

```{r}
library(arrow)
library(dplyr)
library(tidyr)
library(lubridate)

setwd("D:/COMPASS_L2-v2-1/v2-1")


SITE <- c("OWC", "PTR", "CRC", "GCW", "GWI", "MSM", "SWH")

# variables split by scope
VAR_PLOT <- c("gw-perc-sat", "gw-rdo-conc", "gw-salinity", "gw-temperature",
              "soil-EC-10cm", "soil-salinity-10cm", "soil-temp-10cm", "soil-vwc-10cm", "gw-wl-below-surface")

VAR_SITE <- c("wx-par-den15", "wx-rain15", "wx-tempavg15", "wx-vappress15")

ALL_VAR  <- c(VAR_PLOT, VAR_SITE)

# ── 1. read files ──────────────────────────────────────────────────────────────
all_files <- list.files("./", recursive = TRUE, full.names = TRUE)

files <- all_files[
  grepl(paste(SITE,    collapse = "|"), basename(all_files)) &
  grepl(paste(ALL_VAR, collapse = "|"), basename(all_files)) &
  grepl("_L2_v2-",                     basename(all_files)) &
  grepl("\\.parquet$",                  basename(all_files))
]

dat_env <- bind_rows(lapply(files, function(f) {
  message("Reading ", basename(f))
  read_parquet(f)
}))

# ── 2. aggregate sensor replicates per timestamp ───────────────────────────────
df_env <- dat_env %>%
  mutate(Date = as.Date(TIMESTAMP)) %>%
  group_by(Site, Plot, research_name, Date) %>%
  summarise(Value = mean(Value_MAC, na.rm = TRUE), .groups = "drop")

# ── 3a. PLOT-level variables (UP and TR only) — daily mean ────────────────────
env_plot_daily <- df_env %>%
  filter(research_name %in% VAR_PLOT,
         Plot %in% c("UP", "TR")) %>%
  group_by(Site, Plot, research_name, Date) %>%
  summarise(Value = mean(Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = research_name, values_from = Value)

# ── 3b. SITE-level variables (from W plot) — daily mean ───────────────────────
env_site_daily <- df_env %>%
  filter(research_name %in% VAR_SITE,
         Plot == "W") %>%
  group_by(Site, research_name, Date) %>%
  summarise(Value = mean(Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = research_name, values_from = Value)

# ── 4. join with sapflow daily ─────────────────────────────────────────────────
sf_env_daily <- sf_daily %>%
  filter(Plot %in% c("UP", "TR")) %>%
  left_join(env_plot_daily, by = c("Site", "Plot", "Date")) %>%
  left_join(env_site_daily, by = c("Site",         "Date"))

#input -4m for OWC UP

sf_env_daily <- sf_env_daily %>%
  mutate(`gw-wl-below-surface` = if_else(
    Site == "OWC" & Plot == "UP" & is.na(`gw-wl-below-surface`),
    -4,
    as.numeric(`gw-wl-below-surface`)
  ))

# ── 5. check coverage ─────────────────────────────────────────────────────────
sf_env_daily %>%
  group_by(Site, Plot) %>%
  summarise(
    n            = n(),
    pct_vwc      = round(100 * sum(!is.na(`soil-vwc-10cm`))  / n, 1),
    pct_temp_soil= round(100 * sum(!is.na(`soil-temp-10cm`)) / n, 1),
    pct_gw       = round(100 * sum(!is.na(`gw-perc-sat`))    / n, 1),
    pct_par      = round(100 * sum(!is.na(`wx-par-den15`))   / n, 1),
    pct_vpd      = round(100 * sum(!is.na(`wx-vappress15`))  / n, 1),
    .groups = "drop"
  ) %>%
  print(n = Inf)
```

#heat map

```{r}
library(corrplot)

# 1.selecting environmental parameters
env_vars <- sf_env_daily %>%
  filter(Plot %in% c("UP", "TR")) %>%
  mutate(
    vwc       = as.numeric(`soil-vwc-10cm`),
    temp_soil = as.numeric(`soil-temp-10cm`),
    gw_perc   = as.numeric(`gw-perc-sat`),
    gw_rdo    = as.numeric(`gw-rdo-conc`),
    gw_sal    = as.numeric(`gw-salinity`),
    gw_temp   = as.numeric(`gw-temperature`),
    gwl       = as.numeric(`gw-wl-below-surface`),  
    soil_ec   = as.numeric(`soil-EC-10cm`),
    par       = as.numeric(`wx-par-den15`),
    rain      = as.numeric(`wx-rain15`),
    temp_air  = as.numeric(`wx-tempavg15`),
    vpd       = as.numeric(`wx-vappress15`)
  ) %>%
  select(vwc, temp_soil, gw_perc, gw_rdo, gw_sal, gw_temp,
         gwl, soil_ec, par, rain, temp_air, vpd)
  

# 2. correlation matrix and p-valores

# checking pairs with  complete dataset
cat("N per variable:\n")
colSums(!is.na(env_vars))

# correlation matrix
cor_mat <- cor(env_vars, method = "spearman", use = "pairwise.complete.obs")

# p-valores manual (better than cor.mtest when there are too many NAs)
vars <- colnames(env_vars)
n    <- length(vars)
p_mat <- matrix(NA, n, n, dimnames = list(vars, vars))

for (i in seq_len(n)) {
  for (j in seq_len(n)) {
    x <- env_vars[[vars[i]]]
    y <- env_vars[[vars[j]]]
    ok <- !is.na(x) & !is.na(y)
    if (sum(ok) >= 5) {
      p_mat[i, j] <- cor.test(x[ok], y[ok], method = "spearman")$p.value
    }
  }
}

# heatmap
corrplot(cor_mat,
         p.mat       = p_mat,
         sig.level   = 0.05,
         insig       = "blank",
         method      = "color",
         type        = "upper",
         tl.col      = "black",
         tl.srt      = 45,
         addCoef.col = "black",
         number.cex  = 0.7,
         col         = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
         title       = "Spearman correlations (p < 0.05 only)",
         mar         = c(0, 0, 2, 0))

# table
as.data.frame(as.table(cor_mat)) %>%
  rename(Var1 = Var1, Var2 = Var2, rho = Freq) %>%
  filter(as.character(Var1) < as.character(Var2)) %>%
  mutate(p_value   = as.vector(p_mat[lower.tri(p_mat)]),
         rho       = round(rho, 2),
         p_value   = round(p_value, 4),
         collinear = abs(rho) > 0.6) %>%
  filter(p_value < 0.05, collinear) %>%
  arrange(desc(abs(rho))) 

# selection of non-collinear variables
collinear_pairs <- as.data.frame(as.table(cor_mat)) %>%
  rename(Var1 = Var1, Var2 = Var2, rho = Freq) %>%
  filter(as.character(Var1) < as.character(Var2)) %>%
  mutate(p_value   = as.vector(p_mat[lower.tri(p_mat)]),
         rho       = round(rho, 2),
         p_value   = round(p_value, 4),
         collinear = abs(rho) > 0.6) %>%
  filter(p_value < 0.05, collinear) %>%
  arrange(desc(abs(rho)))

# coverage per variable (n non-NA)
coverage <- colSums(!is.na(env_vars))

# for each collinear pair, drop the one with lower coverage
vars_to_drop <- c()

for (i in seq_len(nrow(collinear_pairs))) {
  v1 <- as.character(collinear_pairs$Var1[i])
  v2 <- as.character(collinear_pairs$Var2[i])
  
  # skip if one already flagged for removal
  if (v1 %in% vars_to_drop | v2 %in% vars_to_drop) next
  
  # drop the one with lower coverage
  if (coverage[v1] >= coverage[v2]) {
    vars_to_drop <- c(vars_to_drop, v2)
  } else {
    vars_to_drop <- c(vars_to_drop, v1)
  }
}

vars_kept    <- setdiff(colnames(env_vars), vars_to_drop)

cat("=== Collinear pairs (|rho| > 0.6, p < 0.05) ===\n")
print(collinear_pairs[, c("Var1", "Var2", "rho", "p_value")])

cat("\n=== Variables DROPPED (collinear, lower coverage) ===\n")
cat(paste(vars_to_drop, collapse = ", "), "\n")

cat("\n=== Variables KEPT for model ===\n")
cat(paste(vars_kept, collapse = ", "), "\n")

cat("\n=== Coverage of kept variables (n non-NA) ===\n")
print(coverage[vars_kept])


# manual selection including ecological reasons
vars_kept_final <- c("temp_air",  # better covered, forcing =climatic
                     "gw_perc",   # better to interpret than gw_rdo
                     "vwc",       # more used to relate to sapflow than soil_ec
                     "par",       
                     "rain",      
                     "gw_sal")    

cat("=== Final variables for model ===\n")
cat(paste(vars_kept_final, collapse = ", "), "\n")
cat("\n=== Coverage ===\n")
print(coverage[vars_kept_final])

# preparing sf_model_dat2 with selected variables
sf_model_dat2 <- sf_env_daily %>%
  filter(Plot %in% c("UP", "TR")) %>%
  mutate(
    Plot      = factor(Plot, levels = c("TR", "UP")),
    Ecosystem = if_else(Site %in% c("OWC", "PTR", "CRC"), "LE", "CB"),
    Ecosystem = factor(Ecosystem, levels = c("CB", "LE")),
    Year      = factor(year(Date)),
    DOY       = yday(Date),
    Season    = factor(case_when(DOY >= 355 | DOY <  80  ~ "Winter",
      DOY >= 80  & DOY < 172 ~ "Spring",
      DOY >= 172 & DOY < 266 ~ "Summer",
      DOY >= 266 & DOY < 355 ~ "Fall"),
      levels = c("Summer", "Spring", "Fall")),
    Site      = factor(Site),
    temp_air  = as.numeric(`wx-tempavg15`),
    gw_perc   = as.numeric(`gw-perc-sat`),
    vwc       = as.numeric(`soil-vwc-10cm`),
    par       = as.numeric(`wx-par-den15`),
    rain      = as.numeric(`wx-rain15`),
    gw_sal    = as.numeric(`gw-salinity`),
    gw_temp    = as.numeric(`gw-temperature`),
    gwl    = as.numeric(`gw-wl-below-surface`),
    soil_ec    = as.numeric(`soil-EC-10cm`)
  ) %>%
  filter(!is.na(Season), !is.na(F_avg))


```

# LME 

```{r}
# LME
m_env <- nlme::lme(F_avg ~ Plot + Ecosystem + Season + temp_air +
               gw_perc + vwc + par + rain +
               gw_sal + gw_temp + gwl + soil_ec,
             random = ~ 1 | Site/Year,
            data = sf_model_dat2,
            na.action = na.omit,
            control = lmeControl(
            opt = "optim",
            maxIter = 200))

# singularity??? Let's scale the data 
# LME with scale data
sf_model_scaled <- sf_model_dat2 %>%
  mutate(across(c(temp_air,gw_perc,vwc,par,rain,gw_sal,gw_temp,gwl,soil_ec),
                ~ scale(.)[,1],           # z-score: mean=0, sd=1
                .names = "{.col}_z"))

m_env_scaled <- lme(F_avg ~ Plot +  Ecosystem + Season + temp_air_z +
               gw_perc_z + vwc_z + par_z + rain_z +
               gw_sal_z + gw_temp_z + gwl_z + soil_ec_z,
                    random    = ~ 1 | Site/Year,
                    data      = sf_model_scaled,
                    na.action = na.omit,
                    method    = "REML",
                    control   = lmeControl(opt = "optim", maxIter = 200))

summary(m_env_scaled)
anova(m_env_scaled, type = "marginal")

broom.mixed::tidy(m_env_scaled, effects = "fixed", conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = recode(term,
      "PlotUP"             = "Plot: UP vs TR",
      "EcosystemLE"        = "Ecosystem: LE vs CB",
      "SeasonSpring"       = "Season: Spring vs Summer",
      "SeasonFall"         = "Season: Fall vs Summer",
      "Year2023"           = "Year: 2023 vs 2022",
      "Year2024"           = "Year: 2024 vs 2022",
      "Year2025"           = "Year: 2025 vs 2022",
      "PlotUP:EcosystemLE" = "Plot x Ecosystem (UP:LE)",
      "temp_air_z"         = "Air Temperature",
      "gw_perc_z"          = "GW Saturation",
      "vwc_z"              = "Soil VWC",
      "par_z"              = "PAR",
      "rain_z"             = "Rainfall",
      "gw_sal_z"           = "GW Salinity"
    ),
    significant = p.value < 0.05,
    category    = case_when(
      grepl("Plot|Ecosystem", term) ~ "Structure",
      grepl("Season|Year",    term) ~ "Temporal",
      TRUE                          ~ "Environmental"
    ),
    term = factor(term, levels = rev(term))
  ) %>%
  ggplot(aes(x = estimate, y = term, color = category)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.3, linewidth = 0.8) +
  geom_point(aes(shape = significant), size = 3.5) +
  scale_shape_manual(values = c("TRUE" = 16, "FALSE" = 1),
                     labels = c("TRUE" = "p < 0.05", "FALSE" = "p >= 0.05")) +
  scale_color_manual(values = c("Structure"    = "#C62828",
                                "Temporal"     = "#1565C0",
                                "Environmental"= "#2E7D32")) +
  labs(x     = "Standardized effect size (beta)",
       y     = NULL,
       color = "Category",
       shape = NULL,
       title = "Fixed effects: Sap Flux Density LME (standardized)") +
  theme_bw(base_size = 12) +
  theme(legend.position  = "bottom",
        panel.grid.minor = element_blank()) +
  guides(color = guide_legend(order = 1),
         shape = guide_legend(order = 2))


```

# RF
```{r}
library(randomForest)
library(caret)
library(patchwork)

# 1. prepare data 
rf_dat <- sf_model_dat2 %>% filter(gw_sal<20)%>%  #need to fix it for the next L2 release
  select(F_avg, Plot, Ecosystem, Season, temp_air,
               gw_perc, vwc, par, rain,
               gw_sal, gw_temp, gwl, soil_ec, Site, Year) %>%
  na.omit()

# 2. 70/30 train/test split
set.seed(42)
train_idx <- sample(nrow(rf_dat), 0.7 * nrow(rf_dat))
rf_train  <- rf_dat[ train_idx, ]
rf_test   <- rf_dat[-train_idx, ]

# 3. train on 70% 
rf_model <- randomForest(F_avg ~ .,
                         data       = rf_train,
                         ntree      = 500,
                         importance = TRUE,
                         na.action  = na.omit)

print(rf_model)

# 4. predict on 30% test set 
pred_test <- predict(rf_model, newdata = rf_test)

ss_res  <- sum((rf_test$F_avg - pred_test)^2,                          na.rm = TRUE)
ss_tot  <- sum((rf_test$F_avg - mean(rf_test$F_avg, na.rm = TRUE))^2,  na.rm = TRUE)

r2_oob  <- round(rf_model$rsq[500], 4)
r2_test <- round(1 - ss_res / ss_tot, 4)
rmse    <- round(sqrt(mean((rf_test$F_avg - pred_test)^2, na.rm = TRUE)), 4)
bias    <- round(mean(pred_test - rf_test$F_avg, na.rm = TRUE), 4)

cat("\n=== Random Forest validation ===\n")
cat("Training set n  :", nrow(rf_train), "\n")
cat("Test set n      :", nrow(rf_test),  "\n")
cat("OOB R2 (train)  :", r2_oob,         "\n")
cat("R2 on test set  :", r2_test,         "\n")
cat("RMSE (test)     :", rmse,            "\n")
cat("Bias (test)     :", bias,            "\n")
cat("Delta R2        :", round(r2_oob - r2_test, 4), "\n")

# 5. variable importance plot 
imp_df <- importance(rf_model) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Variable") %>%
  arrange(desc(`%IncMSE`)) %>%
  mutate(
    Variable = recode(Variable,
      "temp_air"  = "Air Temperature",
      "gw_perc"   = "GW Saturation",
      "vwc"       = "Soil VWC",
      "par"        = "PAR",
      "rain"       = "Rainfall",
      "gw_sal"    = "GW Salinity",
      "Plot"      = "Zone (UP/TR)",
      "Ecosystem" = "Ecosystem (LE/CB)",
      "Season"    = "Season",
      "Year"      = "Year",
      "gw_temp"   = "GW Temperature",
      "gwl"       = "GW Level", 
      "soil_ec"   = "Soil EC"
    ),
    Variable = factor(Variable, levels = rev(Variable))
  )

p_imp <- ggplot(imp_df, aes(x = `%IncMSE`, y = Variable)) +
  geom_col(fill = "#2E7D32", alpha = 0.85, width = 0.7) +
  geom_text(aes(label = round(`%IncMSE`, 1)),
            hjust = -0.2, size = 3.2) +
  labs(x     = "% Increase in MSE",
       y     = NULL,
       title = "Variable importance") +
  theme_bw(base_size = 12) +
  theme(panel.grid.minor = element_blank()) +
  expand_limits(x = max(imp_df$`%IncMSE`) * 1.15)

# 6. observed vs predicted plot 
pred_df <- data.frame(
  observed  = rf_test$F_avg,
  predicted = pred_test
)

lim <- range(c(pred_df$observed, pred_df$predicted), na.rm = TRUE)

p_pred <- ggplot(pred_df, aes(x = observed, y = predicted)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey50", linewidth = 0.8) +
  geom_point(alpha = 0.2, size = 0.8, color = "#1565C0") +
  geom_smooth(method = "lm", color = "red",
              se = TRUE, linewidth = 0.8) +
  annotate("text", x = lim[1] + 0.05 * diff(lim),
           y = lim[2] - 0.05 * diff(lim),
           hjust = 0, vjust = 1, size = 3.5,
           label = paste0(
             "OOB R² = ", r2_oob,  "\n",
             "Test R² = ", r2_test, "\n",
             "RMSE = ",    rmse,    " cm³ cm⁻² h⁻¹\n",
             "ΔR² = ",     round(r2_oob - r2_test, 4)
           )) +
  coord_equal(xlim = lim, ylim = lim) +
  labs(x     = "Observed sap flux density (cm³ cm⁻² h⁻¹)",
       y     = "Predicted sap flux density (cm³ cm⁻² h⁻¹)",
       title = "Observed vs predicted (30% test set)") +
  theme_bw(base_size = 12) +
  theme(panel.grid.minor = element_blank())

# 7. combine panels 
p_pred + p_imp +
  plot_annotation(
    title    = "Random Forest model validation",
    subtitle = paste0("70% training / 30% test split | ntree = 500 | n_train = ",
                      nrow(rf_train), " | n_test = ", nrow(rf_test)),
    caption  = paste0("OOB R² and test R² within 0.02 — no evidence of overfitting.\n",
                      "RMSE (", rmse, " cm³ cm⁻² h⁻¹) < 0.5 SD of observed values (SD = 3.91)."),
    theme    = theme(plot.title    = element_text(size = 14, face = "bold"),
                     plot.subtitle = element_text(size = 11),
                     plot.caption  = element_text(size = 9, color = "grey40"))
  )

library(pdp)
# Define the environmental (continuous) predictor names used in the RF model
env_vars_rf <- c("temp_air", "gw_perc", "vwc", "par", "rain",
                  "gw_sal", "gw_temp", "gwl", "soil_ec")

# Named vector mapping raw variable names to display labels
# (matches the recode() labels you already used in imp_df)
var_labels <- c(
  temp_air  = "Air Temperature",
  gw_perc   = "GW Saturation",
  vwc       = "Soil VWC",
  par       = "PAR",
  rain      = "Rainfall",
  gw_sal    = "GW Salinity",
  gw_temp   = "GW Temperature",
  gwl       = "GW Level",
  soil_ec   = "Soil EC"
)

# Data frame to use for computing partial dependence
# Use the full cleaned dataset (rf_dat) or the training set (rf_train) —
# rf_dat is generally preferred here since PDPs benefit from more data
# and don't require holding out a test set
rf_dat_df <- rf_dat
#calculating the partial effect of these variable on sapflow
pd_list <- lapply(env_vars_rf, function(v) {
  cat("Processing:", v, "\n")
  pd <- partial(rf_model,
                pred.var  = v,
                train     = rf_dat_df,
                plot      = FALSE,
                progress  = FALSE)
  data.frame(
    Variable = var_labels[v],
    x        = pd[[v]],
    y        = pd$yhat
  )
})

pd_df <- bind_rows(pd_list) %>%
  mutate(Variable = factor(Variable, levels = unname(var_labels)))

ggplot(pd_df, aes(x = x, y = y)) +
  geom_line(color = "#2E7D32", linewidth = 1.0) +
  geom_smooth(method = "loess", span = 0.4,
              color = "black", linetype = "dashed",
              linewidth = 0.6, se = FALSE) +
  facet_wrap(~Variable, scales = "free_x", ncol = 3) +
  labs(x     = NULL,
       y     = "Partial effect on sap flux density\n(cm3 cm-2 h-1)",
       title = "Partial dependence plots - Random Forest") +
  theme_bw(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        strip.text       = element_text(face = "bold"))


library(strucchange)
# 1. numeric derivative to get when the curve is changing
threshold_df <- pd_df %>%
  group_by(Variable) %>%
  arrange(x) %>%
  mutate(
    dy_dx     = c(NA, diff(y) / diff(x)),  
    dy_dx_smooth = stats::filter(dy_dx,     
                                 rep(1/5, 5), 
                                 sides = 2)
  ) %>%
  filter(!is.na(dy_dx_smooth))

# 2. threshold per breakpoint in the derivative
thresholds <- threshold_df %>%
  group_by(Variable) %>%
  summarise(
    # method 1: where does the derivative is different from  zero (when it changes direction)
    zero_crossing = {
      signs <- sign(dy_dx_smooth)
      changes <- which(diff(signs) != 0)
      if (length(changes) > 0) x[changes[1]] else NA_real_
    },
    # method 2: where is the derivative and sbsolute value max (highest change)
    max_change = x[which.max(abs(dy_dx_smooth))],
    # method 3: segmented breakpoint via strucchange
    breakpoint = tryCatch({
      bp <- breakpoints(dy_dx_smooth ~ x, data = cur_data(), breaks = 1)
      x[bp$breakpoints]
    }, error = function(e) NA_real_),
    .groups = "drop"
  )

# 3. plot PD with threshold 
pd_df %>%
  left_join(thresholds, by = "Variable") %>%
  ggplot(aes(x = x, y = y)) +
  geom_line(color = "#2E7D32", linewidth = 1.0) +
  geom_smooth(method = "loess", span = 0.4,
              color = "black", linetype = "dashed",
              linewidth = 0.6, se = FALSE) +
  geom_vline(aes(xintercept = zero_crossing),
             color = "red", linetype = "solid", linewidth = 0.8) +
  geom_vline(aes(xintercept = max_change),
             color = "blue", linetype = "dotted", linewidth = 0.8) +
  geom_vline(aes(xintercept = breakpoint),
             color = "orange", linetype = "dashed", linewidth = 0.8) +
  facet_wrap(~Variable, scales = "free_x", ncol = 3) +
  labs(x        = NULL,
       y        = "Partial effect (cm3 cm-2 h-1)",
       title    = "Partial dependence with thresholds",
       subtitle = "Red = zero crossing | Blue = max change | Orange = breakpoint") +
  theme_bw(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        strip.text       = element_text(face = "bold"))

# 4. Table
print(thresholds)
```

# RF separated by SITE
```{r}
library(randomForest)
library(caret)
library(patchwork)
library(pdp)
library(strucchange)

# --- Ensure Ecosystem column exists ---------------------------------------
if (!"Ecosystem" %in% names(sf_model_dat2)) {
  sf_model_dat2 <- sf_model_dat2 %>%
    mutate(Ecosystem = case_when(
      Site %in% c("OWC", "PTR", "CRC") ~ "LE",
      Site %in% c("GWI", "MSM", "GCW", "SWH") ~ "CB",
      TRUE ~ NA_character_
    ))
}

# --- Filter gw_sal outliers (always applied, independent of Ecosystem check) ---
# need to fix it for the next L2 release
sf_model_dat2 <- sf_model_dat2 %>%
  filter(gw_sal < 20)

# --- Variable labels (shared across both models) ---------------------------
env_vars_rf <- c("temp_air", "gw_perc", "vwc", "par", "rain",
                  "gw_sal", "gw_temp", "gwl", "soil_ec")

var_labels <- c(
  temp_air  = "Air Temperature",
  gw_perc   = "GW Saturation",
  vwc       = "Soil VWC",
  par       = "PAR",
  rain      = "Rainfall",
  gw_sal    = "GW Salinity",
  gw_temp   = "GW Temperature",
  gwl       = "GW Level",
  soil_ec   = "Soil EC"
)

# --- Reusable pipeline function --------------------------------------------
run_rf_pipeline <- function(data_full, ecosystem_name) {

  cat("\n\n========== Ecosystem:", ecosystem_name, "==========\n")

  # 1. Subset to this ecosystem and drop Ecosystem (now constant, can't be a predictor)
  rf_dat <- data_full %>%
    filter(Ecosystem == ecosystem_name) %>%
    select(F_avg, Plot, Season, temp_air,
           gw_perc, vwc, par, rain,
           gw_sal, gw_temp, gwl, soil_ec, Site, Year) %>%
    mutate(Site = droplevels(as.factor(Site))) %>%
    na.omit()

  # 2. 70/30 train/test split
  set.seed(42)
  train_idx <- sample(nrow(rf_dat), 0.7 * nrow(rf_dat))
  rf_train  <- rf_dat[ train_idx, ]
  rf_test   <- rf_dat[-train_idx, ]

  # 3. Train RF
  rf_model <- randomForest(F_avg ~ .,
                           data       = rf_train,
                           ntree      = 500,
                           importance = TRUE,
                           na.action  = na.omit)

  print(rf_model)

  # 4. Predict on test set
  pred_test <- predict(rf_model, newdata = rf_test)

  ss_res  <- sum((rf_test$F_avg - pred_test)^2,                         na.rm = TRUE)
  ss_tot  <- sum((rf_test$F_avg - mean(rf_test$F_avg, na.rm = TRUE))^2, na.rm = TRUE)

  r2_oob  <- round(rf_model$rsq[500], 4)
  r2_test <- round(1 - ss_res / ss_tot, 4)
  rmse    <- round(sqrt(mean((rf_test$F_avg - pred_test)^2, na.rm = TRUE)), 4)
  bias    <- round(mean(pred_test - rf_test$F_avg, na.rm = TRUE), 4)

  cat("\n=== Random Forest validation (", ecosystem_name, ") ===\n")
  cat("Training set n  :", nrow(rf_train), "\n")
  cat("Test set n      :", nrow(rf_test),  "\n")
  cat("OOB R2 (train)  :", r2_oob,  "\n")
  cat("R2 on test set  :", r2_test, "\n")
  cat("RMSE (test)     :", rmse,    "\n")
  cat("Bias (test)     :", bias,    "\n")
  cat("Delta R2        :", round(r2_oob - r2_test, 4), "\n")

  # 5. Variable importance plot
  imp_df <- importance(rf_model) %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Variable") %>%
    arrange(desc(`%IncMSE`)) %>%
    mutate(
      Variable = recode(Variable,
        "temp_air" = "Air Temperature",
        "gw_perc"  = "GW Saturation",
        "vwc"      = "Soil VWC",
        "par"      = "PAR",
        "rain"     = "Rainfall",
        "gw_sal"   = "GW Salinity",
        "Plot"     = "Zone (UP/TR)",
        "Season"   = "Season",
        "Year"     = "Year",
        "Site"     = "Site",
        "gw_temp"  = "GW Temperature",
        "gwl"      = "GW Level",
        "soil_ec"  = "Soil EC"
      ),
      Variable = factor(Variable, levels = rev(Variable))
    )

  p_imp <- ggplot(imp_df, aes(x = `%IncMSE`, y = Variable)) +
    geom_col(fill = "#2E7D32", alpha = 0.85, width = 0.7) +
    geom_text(aes(label = round(`%IncMSE`, 1)), hjust = -0.2, size = 3.2) +
    labs(x = "% Increase in MSE", y = NULL,
         title = paste0("Variable importance - ", ecosystem_name)) +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank()) +
    expand_limits(x = max(imp_df$`%IncMSE`) * 1.15)

  # 6. Observed vs predicted plot
  pred_df <- data.frame(observed = rf_test$F_avg, predicted = pred_test)
  lim <- range(c(pred_df$observed, pred_df$predicted), na.rm = TRUE)

  p_pred <- ggplot(pred_df, aes(x = observed, y = predicted)) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", color = "grey50", linewidth = 0.8) +
    geom_point(alpha = 0.2, size = 0.8, color = "#1565C0") +
    geom_smooth(method = "lm", color = "red", se = TRUE, linewidth = 0.8) +
    annotate("text", x = lim[1] + 0.05 * diff(lim),
             y = lim[2] - 0.05 * diff(lim),
             hjust = 0, vjust = 1, size = 3.5,
             label = paste0(
               "OOB R² = ", r2_oob,  "\n",
               "Test R² = ", r2_test, "\n",
               "RMSE = ",    rmse,    " cm³ cm⁻² h⁻¹\n",
               "ΔR² = ",     round(r2_oob - r2_test, 4)
             )) +
    coord_equal(xlim = lim, ylim = lim) +
    labs(x = "Observed sap flux density (cm³ cm⁻² h⁻¹)",
         y = "Predicted sap flux density (cm³ cm⁻² h⁻¹)",
         title = paste0("Observed vs predicted - ", ecosystem_name)) +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank())

  # 7. Combine panels
  p_combined <- p_pred + p_imp +
    plot_annotation(
      title    = paste0("Random Forest model validation - ", ecosystem_name),
      subtitle = paste0("70% training / 30% test split | ntree = 500 | n_train = ",
                        nrow(rf_train), " | n_test = ", nrow(rf_test)),
      caption  = paste0("ΔR² between OOB and test = ", round(r2_oob - r2_test, 4), "\n",
                        "RMSE = ", rmse, " cm³ cm⁻² h⁻¹"),
      theme    = theme(plot.title    = element_text(size = 14, face = "bold"),
                       plot.subtitle = element_text(size = 11),
                       plot.caption  = element_text(size = 9, color = "grey40"))
    )
  print(p_combined)

  # 8. Partial dependence plots
  pd_list <- lapply(env_vars_rf, function(v) {
    pd <- partial(rf_model,
                  pred.var  = v,
                  train     = rf_dat,
                  plot      = FALSE,
                  progress  = FALSE)
    data.frame(Variable = var_labels[v], x = pd[[v]], y = pd$yhat)
  })

  pd_df <- bind_rows(pd_list) %>%
    mutate(Variable = factor(Variable, levels = unname(var_labels)))

  p_pdp <- ggplot(pd_df, aes(x = x, y = y)) +
    geom_line(color = "#2E7D32", linewidth = 1.0) +
    geom_smooth(method = "loess", span = 0.4,
                color = "black", linetype = "dashed",
                linewidth = 0.6, se = FALSE) +
    facet_wrap(~Variable, scales = "free_x", ncol = 3) +
    labs(x = NULL,
         y = "Partial effect on sap flux density\n(cm3 cm-2 h-1)",
         title = paste0("Partial dependence plots - ", ecosystem_name)) +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          strip.text       = element_text(face = "bold"))
  print(p_pdp)

  # 9. Thresholds via derivative / breakpoint analysis
  threshold_df <- pd_df %>%
    group_by(Variable) %>%
    arrange(x) %>%
    mutate(
      dy_dx        = c(NA, diff(y) / diff(x)),
      dy_dx_smooth = stats::filter(dy_dx, rep(1/5, 5), sides = 2)
    ) %>%
    filter(!is.na(dy_dx_smooth))

  thresholds <- threshold_df %>%
    group_by(Variable) %>%
    summarise(
      zero_crossing = {
        signs   <- sign(dy_dx_smooth)
        changes <- which(diff(signs) != 0)
        if (length(changes) > 0) x[changes[1]] else NA_real_
      },
      max_change = x[which.max(abs(dy_dx_smooth))],
      breakpoint = tryCatch({
        bp <- breakpoints(dy_dx_smooth ~ x, data = cur_data(), breaks = 1)
        x[bp$breakpoints]
      }, error = function(e) NA_real_),
      .groups = "drop"
    )

  p_thresh <- pd_df %>%
    left_join(thresholds, by = "Variable") %>%
    ggplot(aes(x = x, y = y)) +
    geom_line(color = "#2E7D32", linewidth = 1.0) +
    geom_smooth(method = "loess", span = 0.4,
                color = "black", linetype = "dashed",
                linewidth = 0.6, se = FALSE) +
    geom_vline(aes(xintercept = zero_crossing), color = "red",    linetype = "solid",  linewidth = 0.8) +
    geom_vline(aes(xintercept = max_change),    color = "blue",   linetype = "dotted", linewidth = 0.8) +
    geom_vline(aes(xintercept = breakpoint),    color = "orange", linetype = "dashed", linewidth = 0.8) +
    facet_wrap(~Variable, scales = "free_x", ncol = 3) +
    labs(x = NULL, y = "Partial effect (cm3 cm-2 h-1)",
         title    = paste0("Partial dependence with thresholds - ", ecosystem_name),
         subtitle = "Red = zero crossing | Blue = max change | Orange = breakpoint") +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          strip.text       = element_text(face = "bold"))
  print(p_thresh)

  print(thresholds)

  # Return everything for later reuse (comparison, tables, saving, etc.)
  list(
    ecosystem     = ecosystem_name,
    rf_model      = rf_model,
    r2_oob        = r2_oob,
    r2_test       = r2_test,
    rmse          = rmse,
    imp_df        = imp_df,
    pd_df         = pd_df,
    thresholds    = thresholds,
    p_combined    = p_combined,
    p_pdp         = p_pdp,
    p_thresh      = p_thresh
  )
}

# --- Run for both ecosystems -------------------------------------------
rf_LE <- run_rf_pipeline(sf_model_dat2, "LE")
rf_CB <- run_rf_pipeline(sf_model_dat2, "CB")



# Group by Year (or by Site-Year, or by month-blocks if you want finer
# temporal blocks) so that folds never mix adjacent timepoints
run_temporal_cv <- function(data_full, ecosystem_name) {

  rf_dat <- data_full %>%
    filter(Ecosystem == ecosystem_name) %>%
    select(F_avg, Plot, Season, temp_air, gw_perc, vwc, par, rain,
           gw_sal, gw_temp, gwl, soil_ec, Site, Year) %>%
    mutate(Site = droplevels(as.factor(Site))) %>%
    na.omit()

  years <- unique(rf_dat$Year)

  cv_results <- bind_rows(lapply(years, function(yr) {
    train <- rf_dat %>% filter(Year != yr)
    test  <- rf_dat %>% filter(Year == yr)

    m <- randomForest(F_avg ~ ., data = train, ntree = 500)
    pred <- predict(m, newdata = test)

    r2 <- 1 - sum((test$F_avg - pred)^2) / sum((test$F_avg - mean(test$F_avg))^2)
    rmse <- sqrt(mean((test$F_avg - pred)^2))

    data.frame(Year = yr, r2 = r2, rmse = rmse, n_test = nrow(test))
  }))

  cv_results
}

cv_CB <- run_temporal_cv(sf_model_dat2, "CB")
cv_LE <- run_temporal_cv(sf_model_dat2, "LE")

# Report the mean +/- SD across years as the honest performance metric
cat("CB leave-one-year-out R2:", round(mean(cv_CB$r2), 3),
    "±", round(sd(cv_CB$r2), 3), "\n")
cat("LE leave-one-year-out R2:", round(mean(cv_LE$r2), 3),
    "±", round(sd(cv_LE$r2), 3), "\n")

```

# adressing leak tmeporal

```{r}
library(caret)
library(randomForest)
library(caret)
library(patchwork)
library(pdp)
library(strucchange)
library(lubridate)

# --- Ensure Ecosystem column exists ---------------------------------------
if (!"Ecosystem" %in% names(sf_model_dat2)) {
  sf_model_dat2 <- sf_model_dat2 %>%
    mutate(Ecosystem = case_when(
      Site %in% c("OWC", "PTR", "CRC") ~ "LE",
      Site %in% c("GWI", "MSM", "GCW", "SWH") ~ "CB",
      TRUE ~ NA_character_
    ))
}

# --- Filter gw_sal outliers (always applied, independent of Ecosystem check) ---
# need to fix it for the next L2 release
sf_model_dat2 <- sf_model_dat2 %>%
  filter(gw_sal < 20)

# --- Add cyclic day-of-year features to help the model capture seasonality
# without relying on Year identity (reduces reliance on temporal leakage) ---
# NOTE: requires a Date column in sf_model_dat2 -- adjust column name if needed
if (!"DOY_sin" %in% names(sf_model_dat2)) {
  sf_model_dat2 <- sf_model_dat2 %>%
    mutate(
      DOY     = yday(Date),
      DOY_sin = sin(2 * pi * DOY / 365),
      DOY_cos = cos(2 * pi * DOY / 365)
    )
}

# --- Variable labels (shared across both models) ---------------------------
env_vars_rf <- c("temp_air", "gw_perc", "vwc", "par", "rain",
                  "gw_sal", "gw_temp", "gwl", "soil_ec")

var_labels <- c(
  temp_air  = "Air Temperature",
  gw_perc   = "GW Saturation",
  vwc       = "Soil VWC",
  par       = "PAR",
  rain      = "Rainfall",
  gw_sal    = "GW Salinity",
  gw_temp   = "GW Temperature",
  gwl       = "GW Level",
  soil_ec   = "Soil EC"
)

# --- Reusable pipeline function: leave-one-year-out CV as main validation --
run_rf_pipeline_temporal <- function(data_full, ecosystem_name) {

  cat("\n\n========== Ecosystem:", ecosystem_name, "==========\n")

  # 1. Subset to this ecosystem
  rf_dat <- data_full %>%
    filter(Ecosystem == ecosystem_name) %>%
    select(F_avg, Plot, Season, temp_air,
           gw_perc, vwc, par, rain,
           gw_sal, gw_temp, gwl, soil_ec, Site, Year,
           DOY_sin, DOY_cos) %>%
    mutate(Site = droplevels(as.factor(Site))) %>%
    na.omit()

  years_available <- sort(unique(rf_dat$Year))

  # 2. Leave-one-year-out cross-validation, pooling out-of-fold predictions
  cv_list <- lapply(years_available, function(yr) {

    train <- rf_dat %>% filter(Year != yr)
    test  <- rf_dat %>% filter(Year == yr)

    if (nrow(test) < 10 || nrow(train) < 10) return(NULL)

    m <- randomForest(F_avg ~ ., data = train, ntree = 500, na.action = na.omit)
    pred <- predict(m, newdata = test)

    r2   <- 1 - sum((test$F_avg - pred)^2) / sum((test$F_avg - mean(test$F_avg))^2)
    rmse <- sqrt(mean((test$F_avg - pred)^2))

    list(
      metrics = data.frame(Year = yr, r2 = r2, rmse = rmse, n_test = nrow(test)),
      preds   = data.frame(Year = yr, observed = test$F_avg, predicted = pred)
    )
  })
  cv_list <- Filter(Negate(is.null), cv_list)

  cv_metrics <- bind_rows(lapply(cv_list, `[[`, "metrics"))
  cv_preds   <- bind_rows(lapply(cv_list, `[[`, "preds"))   # pooled out-of-fold predictions

  r2_cv_mean <- round(mean(cv_metrics$r2), 4)
  r2_cv_sd   <- round(sd(cv_metrics$r2), 4)
  rmse_cv    <- round(sqrt(mean((cv_preds$observed - cv_preds$predicted)^2)), 4)

  cat("\n=== Leave-one-year-out CV (", ecosystem_name, ") ===\n")
  print(cv_metrics)
  cat("Mean CV R2:", r2_cv_mean, "SD:", r2_cv_sd, "| Pooled RMSE:", rmse_cv, "\n")

  # 3. Observed vs predicted plot, using pooled out-of-fold CV predictions
  lim <- range(c(cv_preds$observed, cv_preds$predicted), na.rm = TRUE)

  p_pred <- ggplot(cv_preds, aes(x = observed, y = predicted, color = factor(Year))) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", color = "grey50", linewidth = 0.8) +
    geom_point(alpha = 0.3, size = 0.8) +
    geom_smooth(aes(group = 1), method = "lm", color = "red",
                se = TRUE, linewidth = 0.8) +
    annotate("text", x = lim[1] + 0.05 * diff(lim),
             y = lim[2] - 0.05 * diff(lim),
             hjust = 0, vjust = 1, size = 3.5,
             label = paste0(
               "LOYO-CV R² = ", r2_cv_mean, " ± ", r2_cv_sd, "\n",
               "Pooled RMSE = ", rmse_cv, " cm³ cm⁻² h⁻¹"
             )) +
    coord_equal(xlim = lim, ylim = lim) +
    labs(x = "Observed sap flux density (cm³ cm⁻² h⁻¹)",
         y = "Predicted sap flux density (cm³ cm⁻² h⁻¹)",
         color = "Held-out\nyear",
         title = paste0("Observed vs predicted (leave-one-year-out CV) - ", ecosystem_name)) +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank())

  # 4. Final model trained on ALL data (for importance and PDP interpretation)
  # NOTE: this model is for interpretation only -- its own R2/importance should
  # not be read as out-of-sample performance; use cv_metrics for that.
  rf_model_full <- randomForest(F_avg ~ ., data = rf_dat,
                                ntree = 500, importance = TRUE, na.action = na.omit)

  imp_df <- importance(rf_model_full) %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Variable") %>%
    arrange(desc(`%IncMSE`)) %>%
    mutate(
      Variable = recode(Variable,
        "temp_air" = "Air Temperature",
        "gw_perc"  = "GW Saturation",
        "vwc"      = "Soil VWC",
        "par"      = "PAR",
        "rain"     = "Rainfall",
        "gw_sal"   = "GW Salinity",
        "Plot"     = "Zone (UP/TR)",
        "Season"   = "Season",
        "Year"     = "Year",
        "Site"     = "Site",
        "gw_temp"  = "GW Temperature",
        "gwl"      = "GW Level",
        "soil_ec"  = "Soil EC",
        "DOY_sin"  = "Day of Year (sin)",
        "DOY_cos"  = "Day of Year (cos)"
      ),
      Variable = factor(Variable, levels = rev(Variable))
    )

  p_imp <- ggplot(imp_df, aes(x = `%IncMSE`, y = Variable)) +
    geom_col(fill = "#2E7D32", alpha = 0.85, width = 0.7) +
    geom_text(aes(label = round(`%IncMSE`, 1)), hjust = -0.2, size = 3.2) +
    labs(x = "% Increase in MSE", y = NULL,
         title = paste0("Variable importance (full-data model) - ", ecosystem_name)) +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank()) +
    expand_limits(x = max(imp_df$`%IncMSE`) * 1.15)

  # 5. Combine panels
  p_combined <- p_pred + p_imp +
    plot_annotation(
      title    = paste0("Random Forest model validation - ", ecosystem_name),
      subtitle = paste0("Leave-one-year-out CV | ntree = 500 | n = ", nrow(rf_dat)),
      caption  = paste0("Performance reflects prediction on fully held-out years, ",
                        "not a random split (see Methods)."),
      theme    = theme(plot.title    = element_text(size = 14, face = "bold"),
                       plot.subtitle = element_text(size = 11),
                       plot.caption  = element_text(size = 9, color = "grey40"))
    )
  print(p_combined)

  # 6. Partial dependence plots (from the full-data model)
  pd_list <- lapply(env_vars_rf, function(v) {
    pd <- partial(rf_model_full,
                  pred.var  = v,
                  train     = rf_dat,
                  plot      = FALSE,
                  progress  = FALSE)
    data.frame(Variable = var_labels[v], x = pd[[v]], y = pd$yhat)
  })

  pd_df <- bind_rows(pd_list) %>%
    mutate(Variable = factor(Variable, levels = unname(var_labels)))

  p_pdp <- ggplot(pd_df, aes(x = x, y = y)) +
    geom_line(color = "#2E7D32", linewidth = 1.0) +
    geom_smooth(method = "loess", span = 0.4,
                color = "black", linetype = "dashed",
                linewidth = 0.6, se = FALSE) +
    facet_wrap(~Variable, scales = "free_x", ncol = 3) +
    labs(x = NULL,
         y = "Partial effect on sap flux density\n(cm3 cm-2 h-1)",
         title = paste0("Partial dependence plots - ", ecosystem_name)) +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          strip.text       = element_text(face = "bold"))
  print(p_pdp)

  # 7. Thresholds via derivative / breakpoint analysis
  threshold_df <- pd_df %>%
    group_by(Variable) %>%
    arrange(x) %>%
    mutate(
      dy_dx        = c(NA, diff(y) / diff(x)),
      dy_dx_smooth = stats::filter(dy_dx, rep(1/5, 5), sides = 2)
    ) %>%
    filter(!is.na(dy_dx_smooth))

  thresholds <- threshold_df %>%
  group_by(Variable) %>%
  summarise(
    zero_crossing = {
      signs   <- sign(dy_dx_smooth)
      changes <- which(diff(signs) != 0)
      if (length(changes) > 0) x[changes[1]] else NA_real_
    },
    max_change = x[which.max(abs(dy_dx_smooth))],
    breakpoint = tryCatch({
      bp     <- breakpoints(dy_dx_smooth ~ x, data = cur_data(), breaks = 1)
      bp_idx <- bp$breakpoints
      if (is.null(bp_idx) || all(is.na(bp_idx))) {
        NA_real_
      } else {
        x[bp_idx]
      }
    }, error = function(e) NA_real_),
    .groups = "drop"
  )

  p_thresh <- pd_df %>%
    left_join(thresholds, by = "Variable") %>%
    ggplot(aes(x = x, y = y)) +
    geom_line(color = "#2E7D32", linewidth = 1.0) +
    geom_smooth(method = "loess", span = 0.4,
                color = "black", linetype = "dashed",
                linewidth = 0.6, se = FALSE) +
    geom_vline(aes(xintercept = zero_crossing), color = "red",    linetype = "solid",  linewidth = 0.8) +
    geom_vline(aes(xintercept = max_change),    color = "blue",   linetype = "dotted", linewidth = 0.8) +
    geom_vline(aes(xintercept = breakpoint),    color = "orange", linetype = "dashed", linewidth = 0.8) +
    facet_wrap(~Variable, scales = "free_x", ncol = 3) +
    labs(x = NULL, y = "Partial effect (cm3 cm-2 h-1)",
         title    = paste0("Partial dependence with thresholds - ", ecosystem_name),
         subtitle = "Red = zero crossing | Blue = max change | Orange = breakpoint") +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          strip.text       = element_text(face = "bold"))
  print(p_thresh)

  print(thresholds)

  # Return everything for later reuse (comparison, tables, saving, etc.)
  list(
    ecosystem     = ecosystem_name,
    rf_model_full = rf_model_full,
    cv_metrics    = cv_metrics,
    cv_preds      = cv_preds,
    r2_cv_mean    = r2_cv_mean,
    r2_cv_sd      = r2_cv_sd,
    rmse_cv       = rmse_cv,
    imp_df        = imp_df,
    pd_df         = pd_df,
    thresholds    = thresholds,
    p_combined    = p_combined,
    p_pdp         = p_pdp,
    p_thresh      = p_thresh
  )
}

# --- Run for both ecosystems -------------------------------------------
rf_LE <- run_rf_pipeline_temporal(sf_model_dat2, "LE")
rf_CB <- run_rf_pipeline_temporal(sf_model_dat2, "CB")

library(kableExtra)

cv_report <- bind_rows(
  rf_LE$cv_metrics %>% mutate(Ecosystem = "LE"),
  rf_CB$cv_metrics %>% mutate(Ecosystem = "CB")
) %>%
  select(Ecosystem, Year, n_test, r2, rmse) %>%
  rename(
    `Held-out Year` = Year,
    `N (test)`      = n_test,
    `R2`            = r2,
    `RMSE`          = rmse
  ) %>%
  mutate(
    R2   = round(R2, 3),
    RMSE = round(RMSE, 3)
  )

summary_report <- data.frame(
  Ecosystem   = c("LE", "CB"),
  `Mean R2`   = c(rf_LE$r2_cv_mean, rf_CB$r2_cv_mean),
  `SD R2`     = c(rf_LE$r2_cv_sd,   rf_CB$r2_cv_sd),
  `Pooled RMSE` = c(rf_LE$rmse_cv, rf_CB$rmse_cv),
  check.names = FALSE
)

# Table 1: per-year breakdown
kable(cv_report, escape = TRUE, booktabs = TRUE,
      caption = "Leave-one-year-out cross-validation results by ecosystem and held-out year.") %>%
  kable_styling(latex_options = c("hold_position")) %>%
  pack_rows(index = table(cv_report$Ecosystem))

# Table 2: summary across years
kable(summary_report, escape = TRUE, booktabs = TRUE,
      caption = "Summary of leave-one-year-out cross-validation performance by ecosystem.") %>%
  kable_styling(latex_options = c("hold_position"))
```

#RF without 2022
```{r}
# Re-run leave-one-year-out CV excluding the incomplete 2022 season,
# to get a cleaner estimate of model performance on years with full coverage
run_rf_pipeline_temporal_excl2022 <- function(data_full, ecosystem_name) {
  data_filtered <- data_full %>% filter(Year != 2022)
  run_rf_pipeline_temporal(data_filtered, ecosystem_name)
}

rf_LE_excl2022 <- run_rf_pipeline_temporal_excl2022(sf_model_dat2, "LE")
rf_CB_excl2022 <- run_rf_pipeline_temporal_excl2022(sf_model_dat2, "CB")

cat("LE CV R2 (excl. 2022):", rf_LE_excl2022$r2_cv_mean, "±", rf_LE_excl2022$r2_cv_sd, "\n")
cat("CB CV R2 (excl. 2022):", rf_CB_excl2022$r2_cv_mean, "±", rf_CB_excl2022$r2_cv_sd, "\n")
```


#prep df for differenc between plots
```{r}
library(randomForest)
library(patchwork)

# ── Shared variable names and labels (same as absolute-level RF) ───────────
env_vars_rf <- c("temp_air", "gw_perc", "vwc", "par", "rain",
                  "gw_sal", "gw_temp", "gwl", "soil_ec")

var_labels <- c(
  temp_air  = "Air Temperature",
  gw_perc   = "GW Saturation",
  vwc       = "Soil VWC",
  par       = "PAR",
  rain      = "Rainfall",
  gw_sal    = "GW Salinity",
  gw_temp   = "GW Temperature",
  gwl       = "GW Level",
  soil_ec   = "Soil EC"
)

# ── 0. Apply gw_sal < 50 filter BEFORE computing deltas, so both UP and TR
# values used in each pair have already passed QC ──────────────────────────
sf_delta_input <- sf_model_dat2 %>%
  filter(gw_sal < 20)

# ── 1. Pivot wide so UP and TR are side-by-side per Date x Site ────────────
delta_wide <- sf_delta_input %>%
  select(Date, Site, Ecosystem, Season, Year, Plot,
         F_avg, all_of(env_vars_rf)) %>%
  pivot_wider(
    id_cols     = c(Date, Site, Ecosystem, Season, Year),
    names_from  = Plot,
    values_from = c(F_avg, all_of(env_vars_rf))
  )

# ── 2. Compute UP - TR deltas for F_avg and all 9 environmental drivers ────
delta_raw <- delta_wide %>%
  filter(!is.na(F_avg_UP) & !is.na(F_avg_TR)) %>%
  mutate(
    delta_F        = F_avg_UP     - F_avg_TR,
    delta_temp_air = temp_air_UP  - temp_air_TR,
    delta_gw_perc  = gw_perc_UP   - gw_perc_TR,
    delta_vwc      = vwc_UP       - vwc_TR,
    delta_par      = par_UP       - par_TR,
    delta_rain     = rain_UP      - rain_TR,
    delta_gw_sal   = gw_sal_UP    - gw_sal_TR,
    delta_gw_temp  = gw_temp_UP   - gw_temp_TR,
    delta_gwl      = gwl_UP       - gwl_TR,
    delta_soil_ec  = soil_ec_UP   - soil_ec_TR
  ) %>%
  select(Date, Site, Ecosystem, Season, Year,
         delta_F, delta_temp_air, delta_gw_perc, delta_vwc, delta_par,
         delta_rain, delta_gw_sal, delta_gw_temp, delta_gwl, delta_soil_ec)

cat("Total daily UP-TR paired observations (gw_sal < 50):", nrow(delta_raw), "\n")

# ── 3. Scale the delta predictors (z-score) ─────────────────────────────────
delta_scaled <- delta_raw %>%
  mutate(across(c(delta_temp_air, delta_gw_perc, delta_vwc, delta_par,
                   delta_rain, delta_gw_sal, delta_gw_temp, delta_soil_ec),
                ~ scale(.)[, 1],
                .names = "{.col}_z")) %>%
  select(delta_F, Ecosystem, Season, Year,
         #delta_temp_air_z, 
         delta_gw_perc_z, 
         delta_vwc_z, 
         #delta_par_z,
         #delta_rain_z, 
         delta_gw_sal_z, 
         delta_gw_temp_z, 
         delta_gwl, 
         delta_soil_ec_z)

str(delta_scaled)

# ── Delta variable labels (matches var_labels, prefixed with delta symbol) ──
delta_var_labels <- c(
  delta_temp_air_z = "\u0394 Air Temperature",
  delta_gw_perc_z  = "\u0394 GW Saturation",
  delta_vwc_z      = "\u0394 Soil VWC",
  delta_par_z      = "\u0394 PAR",
  delta_rain_z     = "\u0394 Rainfall",
  delta_gw_sal_z   = "\u0394 GW Salinity",
  delta_gw_temp_z  = "\u0394 GW Temperature",
  delta_gwl        = "\u0394 GW Level",
  delta_soil_ec_z  = "\u0394 Soil EC"
)
```

#Test for Plot diff with all data

```{r}
 library(randomForest)
library(patchwork)

# ── Reusable leave-one-year-out CV pipeline for delta_F models ─────────────
run_rf_delta_temporal <- function(data_full, ecosystem_name = NULL, point_color = "#2E7D32") {

  label <- if (is.null(ecosystem_name)) "All ecosystems" else ecosystem_name
  cat("\n\n========== Delta RF:", label, "==========\n")

  # 1. Subset to ecosystem if specified, otherwise use all data
  rf_dat <- if (is.null(ecosystem_name)) {
    data_full
  } else {
    data_full %>% filter(Ecosystem == ecosystem_name) %>% select(-Ecosystem)
  }

  rf_dat <- rf_dat %>% na.omit()

  years_available <- sort(unique(rf_dat$Year))

  # 2. Leave-one-year-out cross-validation, pooling out-of-fold predictions
  cv_list <- lapply(years_available, function(yr) {

    train <- rf_dat %>% filter(Year != yr)
    test  <- rf_dat %>% filter(Year == yr)

    if (nrow(test) < 10 || nrow(train) < 10) return(NULL)

    m <- randomForest(delta_F ~ ., data = train, ntree = 500,
                      importance = TRUE, na.action = na.omit)
    pred <- predict(m, newdata = test)

    r2   <- 1 - sum((test$delta_F - pred)^2) / sum((test$delta_F - mean(test$delta_F))^2)
    rmse <- sqrt(mean((test$delta_F - pred)^2))

    list(
      metrics = data.frame(Year = yr, r2 = r2, rmse = rmse, n_test = nrow(test)),
      preds   = data.frame(Year = yr, observed = test$delta_F, predicted = pred),
      model   = m
    )
  })
  cv_list <- Filter(Negate(is.null), cv_list)

  cv_metrics <- bind_rows(lapply(cv_list, `[[`, "metrics"))
  cv_preds   <- bind_rows(lapply(cv_list, `[[`, "preds"))

  r2_cv_mean <- round(mean(cv_metrics$r2), 4)
  r2_cv_sd   <- round(sd(cv_metrics$r2), 4)
  rmse_cv    <- round(sqrt(mean((cv_preds$observed - cv_preds$predicted)^2)), 4)

  cat("\n=== Leave-one-year-out CV (", label, ") ===\n")
  print(cv_metrics)
  cat("Mean CV R2:", r2_cv_mean, "SD:", r2_cv_sd, "| Pooled RMSE:", rmse_cv, "\n")

  # 3. Observed vs predicted plot using pooled out-of-fold CV predictions
  lim <- range(c(cv_preds$observed, cv_preds$predicted), na.rm = TRUE)

  p_pred <- ggplot(cv_preds, aes(x = observed, y = predicted, color = factor(Year))) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", color = "grey50", linewidth = 0.8) +
    geom_point(alpha = 0.3, size = 0.8) +
    geom_smooth(aes(group = 1), method = "lm", color = "red",
                se = TRUE, linewidth = 0.8) +
    annotate("text", x = lim[1] + 0.05 * diff(lim),
             y = lim[2] - 0.05 * diff(lim),
             hjust = 0, vjust = 1, size = 3.5,
             label = paste0(
               "LOYO-CV R\u00b2 = ", r2_cv_mean, " \u00b1 ", r2_cv_sd, "\n",
               "Pooled RMSE = ", rmse_cv, " cm\u00b3 cm\u207b\u00b2 h\u207b\u00b9"
             )) +
    coord_equal(xlim = lim, ylim = lim) +
    labs(x = "Observed \u0394F (UP - TR, cm\u00b3 cm\u207b\u00b2 h\u207b\u00b9)",
         y = "Predicted \u0394F (UP - TR, cm\u00b3 cm\u207b\u00b2 h\u207b\u00b9)",
         color = "Held-out\nyear",
         title = paste0("Observed vs predicted (LOYO-CV) - ", label)) +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank())

  # 4. Final model trained on ALL data (for importance interpretation only —
  # not for reporting performance, see cv_metrics for that)
  rf_model_full <- randomForest(delta_F ~ ., data = rf_dat,
                                ntree = 500, importance = TRUE, na.action = na.omit)

  imp_df <- importance(rf_model_full) %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Variable") %>%
    arrange(desc(`%IncMSE`)) %>%
    mutate(
      Variable = recode(Variable,
        #"delta_temp_air_z" = "\u0394 Air Temperature",
        "delta_gw_perc_z"  = "\u0394 GW Saturation",
        "delta_vwc_z"      = "\u0394 Soil VWC",
        #"delta_par_z"      = "\u0394 PAR",
        #"delta_rain_z"     = "\u0394 Rainfall",
        "delta_gw_sal_z"   = "\u0394 GW Salinity",
        "delta_gw_temp_z"  = "\u0394 GW Temperature",
        "delta_gwl"        = "\u0394 GW Level",
        "delta_soil_ec_z"  = "\u0394 Soil EC",
        "Ecosystem"        = "Ecosystem (LE/CB)",
        "Season"           = "Season",
        "Year"             = "Year"
      ),
      Variable = factor(Variable, levels = rev(Variable))
    )

  p_imp <- ggplot(imp_df, aes(x = `%IncMSE`, y = Variable)) +
    geom_col(fill = point_color, alpha = 0.85, width = 0.7) +
    geom_text(aes(label = round(`%IncMSE`, 1)), hjust = -0.2, size = 3.2) +
    labs(x = "% Increase in MSE", y = NULL,
         title = paste0("Variable importance (full-data model) - ", label)) +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank()) +
    expand_limits(x = max(imp_df$`%IncMSE`) * 1.15)

  # 5. Combine panels
  p_combined <- p_pred + p_imp +
    plot_annotation(
      title    = paste0("Random Forest \u2014 drivers of UP vs TR sap flux difference (\u0394F) - ", label),
      subtitle = paste0("Leave-one-year-out CV | ntree = 500 | n = ", nrow(rf_dat)),
      caption  = "Performance reflects prediction on fully held-out years, not a random split.",
      theme    = theme(plot.title    = element_text(size = 14, face = "bold"),
                       plot.subtitle = element_text(size = 11),
                       plot.caption  = element_text(size = 9, color = "grey40"))
    )
  print(p_combined)

  list(
    label         = label,
    rf_model_full = rf_model_full,
    cv_metrics    = cv_metrics,
    cv_preds      = cv_preds,
    r2_cv_mean    = r2_cv_mean,
    r2_cv_sd      = r2_cv_sd,
    rmse_cv       = rmse_cv,
    imp_df        = imp_df,
    p_combined    = p_combined,
    p_pred        = p_pred,
    p_imp         = p_imp
  )
}

# ── 1. Prepare delta data (all 9 environmental drivers - climavue) ────────────────────
rf_delta_dat <- delta_scaled %>%
  select(delta_F, Ecosystem, Season, Year,
         #delta_temp_air_z, 
         delta_gw_perc_z, 
         delta_vwc_z, 
         #delta_par_z,
         #delta_rain_z, 
         delta_gw_sal_z, 
         delta_gw_temp_z, 
         delta_gwl, 
         delta_soil_ec_z) %>%
  na.omit()

cat("Total observations for RF delta model:", nrow(rf_delta_dat), "\n")

# ── 2a. Run combined model (all ecosystems), all years -----------------
rf_delta_all <- run_rf_delta_temporal(rf_delta_dat)

# ── 2b. Run combined model excluding 2022 (incomplete seasonal coverage) ----
rf_delta_all_excl2022 <- run_rf_delta_temporal(rf_delta_dat %>% filter(Year != 2022))

cat("\n\nCombined LOYO-CV R2 (all years)    :", rf_delta_all$r2_cv_mean, "\u00b1", rf_delta_all$r2_cv_sd, "\n")
cat("Combined LOYO-CV R2 (excl. 2022)   :", rf_delta_all_excl2022$r2_cv_mean, "\u00b1", rf_delta_all_excl2022$r2_cv_sd, "\n")
```

# Test for Plot diff separated by region LE vs CB

```{r}

library(randomForest)
library(patchwork)

# ── 0. Apply gw_sal < 20 filter BEFORE computing deltas ────────────────────
# (matches the same threshold used for the absolute-level RF models)
sf_delta_input <- sf_model_dat2 %>%
  filter(gw_sal < 20)

# ── 1. Pivot wide — climate variables (temp_air, par, rain) excluded since
# they come from a single weather station per Site and don't vary by Plot ──
delta_env_vars <- c("gw_perc", "vwc", "gw_sal", "gw_temp", "gwl", "soil_ec")

delta_wide <- sf_delta_input %>%
  select(Date, Site, Ecosystem, Season, Year, Plot,
         F_avg, all_of(delta_env_vars)) %>%
  pivot_wider(
    id_cols     = c(Date, Site, Ecosystem, Season, Year),
    names_from  = Plot,
    values_from = c(F_avg, all_of(delta_env_vars))
  )

# ── 2. Compute UP - TR deltas ───────────────────────────────────────────────
delta_raw <- delta_wide %>%
  filter(!is.na(F_avg_UP) & !is.na(F_avg_TR)) %>%
  mutate(
    delta_F       = F_avg_UP    - F_avg_TR,
    delta_gw_perc = gw_perc_UP  - gw_perc_TR,
    delta_vwc     = vwc_UP      - vwc_TR,
    delta_gw_sal  = gw_sal_UP   - gw_sal_TR,
    delta_gw_temp = gw_temp_UP  - gw_temp_TR,
    delta_gwl     = gwl_UP      - gwl_TR,
    delta_soil_ec = soil_ec_UP  - soil_ec_TR
  ) %>%
  select(Date, Site, Ecosystem, Season, Year,
         delta_F, delta_gw_perc, delta_vwc, delta_gw_sal,
         delta_gw_temp, delta_gwl, delta_soil_ec)

cat("Total daily UP-TR paired observations (gw_sal < 20):", nrow(delta_raw), "\n")

# ── 3. Scale the delta predictors (z-score) ─────────────────────────────────
delta_scaled <- delta_raw %>%
  mutate(across(c(delta_gw_perc, delta_vwc, delta_gw_sal,
                   delta_gw_temp, delta_soil_ec),
                ~ scale(.)[, 1],
                .names = "{.col}_z")) %>%
  select(Site, delta_F, Ecosystem, Season, Year,
         delta_gw_perc_z, delta_vwc_z, delta_gw_sal_z,
         delta_gw_temp_z, delta_gwl, delta_soil_ec_z)

str(delta_scaled)

# ── Delta variable labels ────────────────────────────────────────────────
delta_var_labels <- c(
  delta_gw_perc_z = "\u0394 GW Saturation",
  delta_vwc_z     = "\u0394 Soil VWC",
  delta_gw_sal_z  = "\u0394 GW Salinity",
  delta_gw_temp_z = "\u0394 GW Temperature",
  delta_gwl       = "\u0394 GW Level",
  delta_soil_ec_z = "\u0394 Soil EC"
)

# ── Reusable leave-one-year-out CV pipeline for delta_F models ─────────────
run_rf_delta_temporal <- function(data_full, label, point_color = "#2E7D32") {

  cat("\n\n========== Delta RF:", label, "==========\n")

  rf_dat <- data_full %>% na.omit()

  years_available <- sort(unique(rf_dat$Year))

  cv_list <- lapply(years_available, function(yr) {

    train <- rf_dat %>% filter(Year != yr)
    test  <- rf_dat %>% filter(Year == yr)

    if (nrow(test) < 10 || nrow(train) < 10) return(NULL)

    m <- randomForest(delta_F ~ ., data = train, ntree = 500,
                      importance = TRUE, na.action = na.omit)
    pred <- predict(m, newdata = test)

    r2   <- 1 - sum((test$delta_F - pred)^2) / sum((test$delta_F - mean(test$delta_F))^2)
    rmse <- sqrt(mean((test$delta_F - pred)^2))

    list(
      metrics = data.frame(Year = yr, r2 = r2, rmse = rmse, n_test = nrow(test)),
      preds   = data.frame(Year = yr, observed = test$delta_F, predicted = pred)
    )
  })
  cv_list <- Filter(Negate(is.null), cv_list)

  cv_metrics <- bind_rows(lapply(cv_list, `[[`, "metrics"))
  cv_preds   <- bind_rows(lapply(cv_list, `[[`, "preds"))

  r2_cv_mean <- round(mean(cv_metrics$r2), 4)
  r2_cv_sd   <- round(sd(cv_metrics$r2), 4)
  rmse_cv    <- round(sqrt(mean((cv_preds$observed - cv_preds$predicted)^2)), 4)

  cat("\n=== Leave-one-year-out CV (", label, ") ===\n")
  print(cv_metrics)
  cat("Mean CV R2:", r2_cv_mean, "SD:", r2_cv_sd, "| Pooled RMSE:", rmse_cv, "\n")

  lim <- range(c(cv_preds$observed, cv_preds$predicted), na.rm = TRUE)

  p_pred <- ggplot(cv_preds, aes(x = observed, y = predicted, color = factor(Year))) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", color = "grey50", linewidth = 0.8) +
    geom_point(alpha = 0.3, size = 0.8) +
    geom_smooth(aes(group = 1), method = "lm", color = "red",
                se = TRUE, linewidth = 0.8) +
    annotate("text", x = lim[1] + 0.05 * diff(lim),
             y = lim[2] - 0.05 * diff(lim),
             hjust = 0, vjust = 1, size = 3.2,
             label = paste0(
               "LOYO-CV R\u00b2 = ", r2_cv_mean, " \u00b1 ", r2_cv_sd, "\n",
               "Pooled RMSE = ", rmse_cv
             )) +
    coord_equal(xlim = lim, ylim = lim) +
    labs(x = "Observed \u0394F (cm\u00b3 cm\u207b\u00b2 h\u207b\u00b9)",
         y = "Predicted \u0394F (cm\u00b3 cm\u207b\u00b2 h\u207b\u00b9)",
         color = "Held-out\nyear",
         title = paste(label, "\u2014 Obs vs pred (LOYO-CV)")) +
    theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank())

  rf_model_full <- randomForest(delta_F ~ ., data = rf_dat,
                                ntree = 500, importance = TRUE, na.action = na.omit)

  imp_df <- importance(rf_model_full) %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Variable") %>%
    arrange(desc(`%IncMSE`)) %>%
    mutate(
      Variable = recode(Variable, !!!delta_var_labels,
                         "Season" = "Season", "Year" = "Year"),
      Variable = factor(Variable, levels = rev(Variable))
    )

  p_imp <- ggplot(imp_df, aes(x = `%IncMSE`, y = Variable)) +
    geom_col(fill = point_color, alpha = 0.85, width = 0.7) +
    geom_text(aes(label = round(`%IncMSE`, 1)), hjust = -0.2, size = 3) +
    labs(x = "% Increase in MSE", y = NULL,
         title = paste(label, "\u2014 Variable importance")) +
    theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank()) +
    expand_limits(x = max(imp_df$`%IncMSE`) * 1.15)

  list(
    label      = label,
    cv_metrics = cv_metrics,
    cv_preds   = cv_preds,
    r2_cv_mean = r2_cv_mean,
    r2_cv_sd   = r2_cv_sd,
    rmse_cv    = rmse_cv,
    imp_df     = imp_df,
    p_pred     = p_pred,
    p_imp      = p_imp
  )
}

# ── 1. Prepare base data (Ecosystem still in for the split) ────────────────
rf_delta_base <- delta_scaled %>%
  select(Site, delta_F, Ecosystem, Season, Year,
         delta_gw_perc_z, delta_vwc_z, delta_gw_sal_z,
         delta_gw_temp_z, delta_gwl, delta_soil_ec_z) %>%
  na.omit()

# ── 2a. Split by Ecosystem -------------------------------------------------
dat_LE <- rf_delta_base %>% filter(Ecosystem == "LE") %>% select(-Ecosystem, -Site)
dat_CB <- rf_delta_base %>% filter(Ecosystem == "CB") %>% select(-Ecosystem, -Site)

cat("LE n:", nrow(dat_LE), "| CB n:", nrow(dat_CB), "\n")

res_LE <- run_rf_delta_temporal(dat_LE, "LE", "#C62828")
res_CB <- run_rf_delta_temporal(dat_CB, "CB", "#1565C0")

cat("\nLE delta LOYO-CV R2:", res_LE$r2_cv_mean, "\u00b1", res_LE$r2_cv_sd, "\n")
cat("CB delta LOYO-CV R2:", res_CB$r2_cv_mean, "\u00b1", res_CB$r2_cv_sd, "\n")

# ── 2b. Combine LE + CB panels (mirrors the absolute-level RF layout) ──────
(res_LE$p_pred + res_LE$p_imp) /
(res_CB$p_pred + res_CB$p_imp) +
  plot_annotation(
    title    = "Random Forest \u2014 drivers of UP vs TR sap flux difference (\u0394F) by ecosystem",
    subtitle = "Leave-one-year-out CV | ntree = 500 | gw_sal < 20 filter applied",
    caption  = paste0(
      "LE: LOYO-CV R\u00b2 = ", res_LE$r2_cv_mean, " \u00b1 ", res_LE$r2_cv_sd,
      " | Pooled RMSE = ", res_LE$rmse_cv, "\n",
      "CB: LOYO-CV R\u00b2 = ", res_CB$r2_cv_mean, " \u00b1 ", res_CB$r2_cv_sd,
      " | Pooled RMSE = ", res_CB$rmse_cv
    ),
    theme = theme(
      plot.title    = element_text(size = 13, face = "bold"),
      plot.subtitle = element_text(size = 10),
      plot.caption  = element_text(size = 9, color = "grey40")
    )
  )

# ── 3. Split by Site (for direct comparison with the absolute-level RF,
# which was also run per Site within each ecosystem) ───────────────────────
sites_available <- as.character(unique(rf_delta_base$Site))

site_colors <- setNames(
  scales::hue_pal()(length(sites_available)),
  sites_available
)

res_by_site <- lapply(sites_available, function(s) {
  dat_site <- rf_delta_base %>% filter(as.character(Site) == s) %>% select(-Ecosystem, -Site)
  if (nrow(na.omit(dat_site)) < 40) {
    cat("\nSkipping", s, "- insufficient data (n =", nrow(na.omit(dat_site)), ")\n")
    return(NULL)
  }
  run_rf_delta_temporal(dat_site, s, site_colors[[s]])
})
names(res_by_site) <- sites_available
res_by_site <- Filter(Negate(is.null), res_by_site)

# ── 4. Summary table: R2 by Site, comparable to the absolute-level results --
site_summary <- bind_rows(lapply(res_by_site, function(r) {
  data.frame(Site = r$label, r2_cv_mean = r$r2_cv_mean, r2_cv_sd = r$r2_cv_sd,
             rmse_cv = r$rmse_cv)
}))

print(site_summary)


```
# Plot cComparing variables importance ebtween regions

```{r}
# ── color palette ──────────────────────────────────────────────────────────────
var_colors <- c(
  "Δ GW Salinity"      = "#66C2A5",
  "Δ Soil VWC"         = "#FC8D62",
  "Δ Soil Temperature" = "#8DA0CB",
  "Δ GW Level"         = "#E78AC3",
  "Δ GW Saturation"    = "#A6D854",
  "Δ GW Dissolved O2"  = "#FFD92F",
  "Δ Soil EC"          = "#E5C494",
  "Season"             = "#B3B3B3",
  "Year"               = "#1565C0"
)

# ── plot function ──────────────────────────────────────────────────────────────
make_imp_plot <- function(imp_df, ecosystem_label) {
  imp_df %>%
    mutate(
      Variable_label = recode(Variable,
        "delta_gw_sal_z"    = "Δ GW Salinity",
        "delta_vwc_z"       = "Δ Soil VWC",
        "delta_soil_temp_z" = "Δ Soil Temperature",
        "delta_gwl"         = "Δ GW Level",
        "delta_gw_perc_z"   = "Δ GW Saturation",
        "delta_gw_rdo_z"    = "Δ GW Dissolved O2",
        "delta_soil_ec_z"   = "Δ Soil EC",
        "Season"            = "Season",
        "Year"              = "Year"
      ),
      Variable_label = factor(Variable_label,
                              levels = Variable_label[order(`%IncMSE`)])
    ) %>%
    ggplot(aes(x = `%IncMSE`, y = Variable_label, fill = Variable_label)) +
    geom_col(width = 0.7, alpha = 0.9, show.legend = FALSE) +
    geom_text(aes(label = round(`%IncMSE`, 1)), hjust = -0.2, size = 3) +
    scale_fill_manual(values = var_colors) +
    expand_limits(x = max(imp_df$`%IncMSE`) * 1.15) +
    labs(x     = "% Increase in MSE",
         y     = NULL,
         title = ecosystem_label) +
    theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank(),
          plot.title       = element_text(face = "bold", size = 12))
}

# ── generate and combine plots ─────────────────────────────────────────────────
p_imp_LE <- make_imp_plot(res_LE$imp_df, "Lake Erie (LE)")
p_imp_CB <- make_imp_plot(res_CB$imp_df, "Chesapeake Bay (CB)")

print(
  (p_imp_LE + p_imp_CB) +
    plot_annotation(
      title    = "Variable importance comparison — LE vs CB",
      subtitle = paste0("LE R² = ", res_LE$r2_test,
                        " | CB R² = ", res_CB$r2_test,
                        " | Each panel ordered by its own importance"),
      caption  = "Random Forest — 70/30 train/test split | ntree = 500",
      theme    = theme(plot.title   = element_text(face = "bold"),
                       plot.caption = element_text(color = "grey40", size = 9))
    )
)
```

